Evaluating the Performance and Energy Efficiency of the \textsc{COSMO-ART} Model System

### 2014-05-29 trunk/paper_enahpc14/Sec2:COSMO-ART.tex
Manuel's changes + rebuttal letter

The   considered  \cosmoart   model  system   is  configured   with  a
semi-Lagrangian   horizontal    advection   sche\-me   with   tricubic
interpolation  and selective filling  diffusion option  in combination
with   the   dynamical    core   using   Runge-Kutta   time   stepping
\cite{COSMO-PartI-2011}.    It   also  makes   use   of  the   Kinetic
PreProcessor solver  (KPP) for  the solution of  atmospheric chemistry
ordinary  differential equations  \cite{Damian-2002}.   Concerning the
modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
parameterizations,   precipitation  formation   is   performed  by   a
two-moment cloud microphysics.

>>>>

The   considered  \cosmoart   model  system   is  configured   with  a
semi-Lagrangian   horizontal    advection   sche\-me   with   tricubic
interpolation  and selective filling  diffusion option  in combination
with   the   dynamical    core   using   Runge-Kutta   time   stepping
\cite{COSMO-PartI-2011}.    It   also  makes   use   of  the   Kinetic
PreProcessor solver  (KPP) for  the solution of  atmospheric chemistry
ordinary  differential equations  \cite{Damian-2002}.   Concerning the
modeling  of  wet  deposition  in  aerosols, the  baseline  has  only
indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
parameterizations,   precipitation  formation   is   performed  by   a
two-moment cloud microphysics.

### 2014-05-29 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes + rebuttal letter

For the experiments with \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings. Specifically, the OpenMPI 1.6.5 library, 
features two operation modes, blocking and polling, which can be selected before \cosmoart
is launched. In the polling configuration (the default mode) ``MPI'' continually polls
the network interface to check for the completion of an event (e.g. send or receive).
This usually yields low latency but high CPU utilization. Thus, one can expect that this mode 
attains the best performance, possibly at the cost a higher energy usage. In the blocking
mode, the CPU is yield to other processes/threads if there are no incoming messages.

\subsection{Time-power-energy analysis of \cosmoart}
\label{subsec:4.2}

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
    \caption{\monch: Isola E1 Rack 2 Total Power.}
    \label{fig:1}
  \end{center}
\end{figure}

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
    \caption{\pilat: Isola HD Total Power.}
    \label{fig:2}
  \end{center}
\end{figure}

Figures~\ref{fig:1}   and   \ref{fig:2}   account   respectively   for
\monch's Isola E1 Rack  2 and \pilat' Isola HD total
power measurements for  1-day or 2-day simulations. On  the Intel Ivy
Bridge EP  based cluster (i.e.  \monch),  the 1-day simulation
was issued  only twice due  to usage restrictions. As  time resolution
was set to one update every  5 minutes for power sampling, the of average
power dissipation was computed by considering 6 values for each single
run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
1-day simulation  was issued  four times and  a 2-day run  only once.
Similarly, the average of power dissipation was computed by considering 4
values  for  each  single  1-day  run  and 9  values  for  the  2-days
run. Corresponding results are gathered in Table~\ref{tab:3}.

\begin{table}[htbf]
  \begin{center}
    \caption{Average power dissipation (W) of the platforms}
    \label{tab:3}
    \begin{tabular}{cccc}
      \hline\noalign{\smallskip}
      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}}\\
      & & \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}} \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      18035.0 & 12622.5 & 3713.6 & 3651.8 \\ 
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{center}
\end{table}

In  Figure~\ref{fig:3}, we  compare both  TTS (right  Y-axis)  and ETS
(left  Y-axis)  metrics  on  both  platforms.  As  expected,  Xeon  E5
outperforms Ivy Bridge EP, being  roughly 130\,\% faster.  The reason for
that is  two-fold: \emph{i}) it has  higher clock frequency  than Ivy Bridge
(2.6\,GHz  against  2.2\,GHz),  and  \emph{ii})  it  aims at  computing  speed
regardless to  energy consumption.  In our experiments,  Ivy Bridge EP
showed the best energy-to-solution, reducing the energy consumption of
Xeon E5 by approximately 7\,\%.

Averaged TTS and ETS results over 10 repetitions of 24h simulation on \tinto are shown in Figure~\ref{fig:4}. Correponding results for average total power are shown in Table~\ref{tab:3}.  
As expected, the average total power dissipated by \cosmoart while using the blocking MPI policy is about 2\,\% less than that consumed by the polling policy. This is due to the power-friendly (blocking) waits performed by the CPU cores while using the blocking MPI mode, in contrast to the polling waits that take place when the polling MPI mode is set. From the point of view of the execution time, there is a slight reduction for the blocking MPI mode. One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the total execution time, however, applications with extensive computing, as e.g. \cosmoart, may show better performance with the blocking modes. This effect works in favor with our results since the energy consumed is also reduced (2\,\%) just by changing the behavior of the OpenMPI library to the blocking policy, i.e, with the \texttt{--mca mpi\_yield\_when\_idle} parameter set in the \texttt{mpirun} command.

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
  \caption{Time-to-solution and energy-to-solution comparison between
    Xeon E5 and Ivy Bridge-EP architectures for a 24h simulation.}
  \label{fig:3}
\end{figure}

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-5.eps}
  \caption{Mean time-to-solution and energy-to-solution on
    \tinto for a 24h simulation using the polling and blocking MPI policies. The 95\,\% confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
    an improvement for blocking policy in both TTS and ETS.}
  \label{fig:4}
\end{figure}

\subsection{Power-performance tracing on \tinto}
\label{subsec:4.3}

A full tracing experiment is conducted on \tinto in order to capture an overall power
profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
for a full-day simulation and analyze the power consumption. Second we correlate performance 
with power traces of two 1-hour time frames during the day: mid-day (11h--12h) and
midnight (23h--24h). All the experiments were performed using 192 \cosmoart processes
m 12 processes in each of node.

% For all the experiments we also analyze the contribution of the MPI library to the energy 
% consumption and how the use of the blocking versus polling message-passing policies can 
% potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
% features two operation modes, blocking and modes, which can be selected before \cosmoart is 
% launched. In the polling configuration (the default mode) ``MPI'' continually polls
% the network interface to check for the completion of an event (e.g. send or receive).
% This usually yields low latency but high CPU utilization. Thus, can one expect that this mode attains the best performance, possibly at the cost a higher energy usage. In the blocking
% mode, the CPU is yield to other processes/threads if there are no incoming messages.

In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a full-day simulation of \cosmoart. These profiles, obtained with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}. The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours. We relate these variations due to the chemical reactions and aerosols computed by ART, that are taking place during daylight hours but not during night. In this case peak power consumption is about 3821\,W. Looking at the power profile using the blocking MPI policy, we observe that this pattern is repeated, favoring the reduction of power during night hours in a more noticeable way than that for the polling case.The peak power dissipation during day hours in this mode is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode. Also, we notice systematic power drops in each hour simulated. We relate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage during. Our final observation is the reduction of the TTS in a factor of 2\,\% with the blocking MPI policy with respect to the polling one.

To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces so that they allow us to correlate the type of tasks being executed with the total power dissipated at each time (see Figure~\ref{fig:11}). For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a full-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB). To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and during midnight (from 23h to 24h), in which \cosmoart perform different internal computations due to the different chemical reactions that happen during day and night. At the same time we also leverage the two already stated MPI policies: polling and blocking.

All the traces in Figure~\ref{fig:11} are divided in different phases. In first phase all the processes perform group communications and execute MPI-receive primitives (e.g. \texttt{MPI\_Recv}). The second performs computation bursts, in between group communications, blocking sends, immediate receives and MPI-wait primitives are executed. Finally, group communications interlaced with computation and a significant synchronization phase is performed. From the performance point of view, one can notice a slight reduction of the TTS that blocking MPI mode leads over the polling MPI policy (around 2\,\%). A reduction of power of about  using the blocking MPI policy is produced only in routines that actually perfom synchronizations, group communications and MPI-receive/-wait primitives. Thus, the reduction of the pair power-energy directly depends on the percentage of time in which blocking MPI operations are executed. For our 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked. Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute the reduction of power, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (2\,\%).

\begin{figure*}[ht]
  \centering
  \hspace{0.8cm}
  \scalebox{0.55}{\input{Figs/24hour.pstex_t}}
  \caption{24 hours simulation trace using the MPI blocking mode on \tinto.}
  \label{fig:9}
\end{figure*}

% \begin{figure*}[htbf]
%   \centering
%   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
%   Percentage of time for the 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
%   \caption{blabla}
%   \label{fig:8}
% \end{figure*}

\begin{figure*}[ht]
  \centering
  \scriptsize
  Simulation of 1 hour during midday (11h--12h) leveraging the polling MPI policy\\
  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
  Simulation of 1 hour during midday (11h--12h) leveraging the blocking MPI policy\\
  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
  1 hour simulation during midnight (23h--24h) leveraging the polling MPI policy\\
  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
  1 hour simulation during midnight (23h--24h) leveraging the blocking MPI policy\\
  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\
  \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
& &  & &  & &  \\[-0.15cm]
Computation     & \multicolumn{1}{>{\columncolor[RGB]{  0,  , 255}}p{0.4cm}}{}  & & 
Block. send     & \multicolumn{1}{>{\columncolor[RGB]{255,  0,174}}p{0.4cm}}{}  & & 
Synchronization & \multicolumn{1}{>{\columncolor[RGB]{179,  0,  0}}p{0.4cm}}{}  & & 
Wait/Wait all   & \multicolumn{1}{>{\columncolor[RGB]{235,255,  0}}p{0.4cm}}{}  \\
& &  & &  & &  \\[-0.15cm]
Immediate recv. & \multicolumn{1}{>{\columncolor[RGB]{100,100,177}}p{0.4cm}}{}  & &
Waiting a msg.  & \multicolumn{1}{>{\columncolor[RGB]{255,  0  ,0}}p{0.4cm}}{}  & & 
Group comm.     & \multicolumn{1}{>{\columncolor[RGB]{255,144, 26}}p{0.4cm}}{}  & &
Others          & \multicolumn{1}{>{\columncolor[RGB]{192,224,  0}}p{0.4cm}}{}  \\
%I/O             & \multicolumn{1}{>{\columncolor[RGB]{172,174, 41}}p{0.4cm}}{}  & &
%
%Idle            & \multicolumn{1}{>{\columncolor[RGB]{117,195,255}}p{0.4cm}}{}  \\
\end{tabular}
  \caption{Power-performance traces during midday and midnight leveraging the polling and blocking MPI policies on \tinto.}
  \label{fig:11}
\end{figure*}

>>>>

For  the   experiments  on  \tinto  platform,  we   also  analyze  the
contribution of the MPI library  to the energy consumption and how the
use  of  the aggressive and degraded message-passing progression engine 
policies can potentially  render energy  savings.

Specifically,   the  OpenMPI 1.6.5  library,  features  two  operation 
policies, \emph{aggressive} and \emph{degraded}, which can be  selected 
before \cosmoart is launched. In the \emph{aggressive} (default) policy,  
normally used in exactly- or under-subscribed mode  (i.e., the  number 
of running processes is equal to or less than the number of  available 
cores), MPI processes will never voluntarily give up the processor  to 
other processes. With some network transports, OpenMPI processes  will 
spin  in  tight  loops  attempting  to  make message passing progress, 
effectively causing other processes to not get  any  CPU  cycles  (and  
therefore never make any  progress).  This  usually  yields  the  best 
performance but high CPU utilization, possibly at the cost of a higher 
energy usage.  On the other hand, the \emph{degraded} policy, suitable 
for the oversubscribed mode (i.e., more  processes  are  running  than 
total cores  available),  MPI  processes  will  frequently  yield  the 
processor to its peers, thereby allowing all processes to make progress.

In our experiments with \cosmoart only the exactly-subscribed mode  is
used. Therefore, \cosmoart processes running  in  the  \emph{degraded} 
policy are constantly checking for the completion  of  an  event (e.g. 
send or receive) and repeatedly calling to  \texttt{sched\_yield()} to 
be picked again by the Linux scheduler. Although one can expect no benefits 
from   using   the  \emph{degraded}   policy   compared   with     the 
\emph{aggressive} one, notable differences can be observed in the CPU utilization. 
In a separated  test  with  the  \emph{aggressive}  policy 
enabled, we noticed that a blocked MPI process waiting for an incoming 
message has CPU utilization of 100\,\%,  however  when  the \emph{degraded} 
mode  is  set,  the  CPU  utilization is only $50\,\%$, approximately. 
Although neither of both policies avoid busy-waiting loops in blocking 
routines, the use of the \emph{degraded} policy is as much as we can 
do in OpenMPI to reduce the CPU utilization, and thus save energy.

% polling configuration
% (the default mode) ``MPI''  continually polls the network interface to
% check for  the completion  of an event  (e.g. send or  receive).  This
% usually yields  low latency  but high CPU  utilization. Thus,  one can
% expect that  this mode attains  the best performance, possibly  at the
% cost a higher energy usage. In  the blocking mode, the CPU is yield to
% other     processes/threads    if     there     are    no     incoming
% messages~\cite{Castillo-2012}.

\subsection{Time-power-energy analysis}
\label{subsec:4.2}

Figures~\ref{fig:1} and \ref{fig:2}  account respectively for \monch's
Isola  E1 Rack 2  and \pilat'  Isola HD  total power  measurements for
1-day or 2-day simulations. On \monch, the 1-day simulation was issued
only twice due  to usage restrictions.  As time  resolution was set to
one  update every  5 minutes  for power  sampling, the  averaged power
consumption was computed by considering  6 values for each single run.
On \pilat, the 1-day simulation was  issued four times and a 2-day run
only once.  Similarly, the  averaged power consumption was computed by
considering 4  values for each single  1-day run and 9  values for the
2-days run. Corresponding results are gathered in Table~\ref{tab:1}.

\begin{figure}[htbf]
  \centering
  \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
  \caption{Isola E1 Rack 2 total power of \monch.}
  \label{fig:1}
\end{figure}

\begin{figure}[htbf]
  \centering
  \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
  \caption{Isola HD total power consumption of \pilat.}
  \label{fig:2}
\end{figure}

\begin{table}[htbf]
  \centering
  \caption{Averaged power consumption (W) of the platforms.}
  \label{tab:1}
  \begin{tabular}{cccc}
    \hline\noalign{\smallskip}     \textbf{\scriptsize{PILATUS}}     &
    \textbf{\scriptsize{MONCH}}  &  \textbf{\scriptsize{TINTORRUM}}  &
    \textbf{\scriptsize{TINTORRUM}}           \\          &          &
    \textbf{\scriptsize{(\emph{Aggressive})}} & \textbf{\scriptsize{(\emph{Degraded})}}
    \\    \noalign{\smallskip}\hline\noalign{\smallskip}   18035.0   &
    12622.5 & 3713.6 & 3651.8 \\ \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

In  Figure~\ref{fig:3}, we  compare both  TTS (right  Y-axis)  and ETS
(left  Y-axis)  metrics  on  both  platforms.  As  expected, Intel Sandy Bridge EP processors of \pilat
outperform the Intel Ivy  Bridge EP processors of \monch, being roughly 130\,\%  faster.  The reason
for that is two-fold: \emph{i}) they have higher clock frequency than Ivy
Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii})  they  aim  at
computing   speed  regardless   to  be   energy  efficient.    In  our
experiments,  \monch  showed  the  best  energy-to-solution,
reducing the energy consumption of \pilat by approximately 7\,\%.

Averaged TTS and ETS results  over 10 repetitions of 24h simulation on
\tinto are shown in  Figure~\ref{fig:4}. The corresponding results for
the  averaged total  power are  also shown  in  Table~\ref{tab:1}.  As
expected, the average total power consumption by \cosmoart while using
the \emph{degraded} MPI policy is about  2\,\% less than that consumed by the
\emph{aggressive} one. This is  due to  the reduction of CPU usage (of about 50\,\%) while using  the \emph{degraded} MPI policy, in
contrast with the \emph{aggressive} policy that always yields to 100\,\% of CPU utilization while waiting for an incoming message. From the point of  view of the execution time, there is a
slight reduction  for the \emph{degraded} MPI  policy. One can  expect that the
costs of the context switches produced in the \emph{degraded} MPI policy can lead to
an  increase  of  the   execution  time,  however,  applications  with
extensive computing,  as e.g.  \cosmoart, may show  better performance
with the \emph{degraded} policy. This effect works in  favor with our results
since the energy consumed is also reduced (2\,\%) just by changing the
behavior of the OpenMPI library  to the \emph{degraded} policy, i.e, with the
\texttt{--mca    mpi\_yield\_when\_idle 1}   parameter   set    in   the
\texttt{mpirun} command.

\begin{figure}[htbf]
  \centering
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
  \caption{Time-to-solution and  energy-to-solution comparison between
    \pilat and \monch for a 24h simulation.}
  \label{fig:3}
\end{figure}

\begin{figure}[htbf]
  \centering
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
  \vspace{-1cm}
  \caption{Mean time-to-solution and  energy-to-solution on \tinto for
    a 24h simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
    95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
    samples) clearly  indicates an improvement for the \emph{degraded} policy in
    both TTS and ETS.}
  \label{fig:4}
\end{figure}

\begin{figure*}[ht]
  \centering
  \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
  \caption{24  hours  simulation power  trace  using  the \emph{aggressive}  and
    \emph{degraded} MPI policies on \tinto.}
  \label{fig:5}
\end{figure*}

\begin{figure*}[ht]
  \centering
  \scriptsize
  Simulation of 1 hour during midday (11h--12h) using the \emph{aggressive} MPI policy \\
  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}} \\
  Simulation of 1 hour during midday (11h--12h) using the \emph{degraded} MPI policy \\
  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}} \\
  1 hour simulation during midnight (23h--24h) using the \emph{aggressive} MPI policy \\
  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}} \\
  1 hour simulation during midnight (23h--24h) using the \emph{degraded} MPI policy \\
  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}} \\
  \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
    & &  & &  & &  \\[-0.15cm]
    Computation     & \multicolumn{1}{>{\columncolor[RGB]{  0,  , 255}}p{0.4cm}}{} & & 
    Block. send     & \multicolumn{1}{>{\columncolor[RGB]{255,  0,174}}p{0.4cm}}{} & & 
    Synchronization & \multicolumn{1}{>{\columncolor[RGB]{179,  0,  0}}p{0.4cm}}{} & & 
    Wait/Wait all   & \multicolumn{1}{>{\columncolor[RGB]{235,255,  0}}p{0.4cm}}{} \\
    & &  & &  & &  \\[-0.15cm]
    Immediate recv. & \multicolumn{1}{>{\columncolor[RGB]{100,100,177}}p{0.4cm}}{} & &
    Waiting a msg.  & \multicolumn{1}{>{\columncolor[RGB]{255,  0  ,0}}p{0.4cm}}{} & & 
    Group comm.     & \multicolumn{1}{>{\columncolor[RGB]{255,144, 26}}p{0.4cm}}{} & &
    Others          & \multicolumn{1}{>{\columncolor[RGB]{192,224,  0}}p{0.4cm}}{} \\
    %I/O             & \multicolumn{1}{>{\columncolor[RGB]{172,174, 41}}p{0.4cm}}{} & &
    %
    %Idle            & \multicolumn{1}{>{\columncolor[RGB]{117,195,255}}p{0.4cm}}{} \\
  \end{tabular}
  \caption{Power-performance   traces  during   midday   and  midnight
    leveraging the \emph{aggressive} and \emph{degraded} MPI policies on \tinto.}
  \label{fig:6}
\end{figure*}

\subsection{Power-performance profiling and tracing}
\label{subsec:4.3}

A full tracing  experiment is conducted on \tinto  in order to capture
an overall power profile at a much finer resolution using the attached
PDUs in each  node. First we run \cosmoart for  a 1-day simulation and
analyze the  power consumption.  Second we  correlate performance with
power  traces  of  two  1-hour  time frames  during  the  day:  midday
(11h--12h) and midnight (23h--24h). All the experiments on \tinto were performed
using 192 cores with 12 MPI processes per node, i.e., using an exactly-subscribed MPI mode.

In order to obtain an initial overview of the power consumption of the
system  model,  we  obtained  power  profiles  leveraging  our  \pmlib
power-measurement    framework   over    a    1-day   simulation    of
\cosmoart. These profiles, produced with both \emph{aggressive} and \emph{degraded} MPI
policies,  are depicted in  Figure~\ref{fig:5}. The  first observation
for the \emph{aggressive} MPI policy is that the first and last hours of the day
consume slightly less  power than the midday hours.  We relate these
variations  due   to  the  different   chemical  reactions,  aerosols,
radiations  and clouds  formation processes,  computed by  ART, taking
place during  daylight hours  but not during  night. In this  case the
peak power consumption is about  3821\,W. Looking at the power profile
using the \emph{degraded}  MPI policy, we observe that  the 1-hour pattern is
repeated along  the 24-hours, favoring  the reduction of  power during
night  hours in  a  more noticeable  way  than that  with the   \emph{aggressive}
policy.   The peak  power consumption  during day  hours  while the
 \emph{degraded}  MPI policy  is leveraged  is  about 3768\,W,  which means  a
reduction of around  50\,W with respect to the  \emph{aggressive} mode. Also, we
notice systematic  power drops for each simulated  hour.  We associate
them  to periods  in  which  the processes  were waiting  for
incoming messages  or synchronizations, thus  yielding the CPU  to a lower
usage. Our final observation is the reduction of the TTS by 2\,\% with
respect to the  \emph{degraded} MPI policy.

To  gain  more  insights   about  the  power-performance  behavior  of
\cosmoart,  we  obtained power-per\-for\-man\-ce  traces  to allow  us
correlate  the type  of  tasks  being executed  with  the total  power
consumption  along  the   time  (see  Figure~\ref{fig:6}).   For  this
experiment we  limit the simulation  time to 1 hour  because: \emph{i})
the  pattern of  the computation  performed by  \cosmoart  is repeated
hourly and, \emph{ii}) the weight  of a 1-day simulation trace can not
be easily handled using our  machines and memory available (each trace
file was about 45\,GB). Thus, we shrink the simulations
to 1 hour during midday (from 11h  to 12h) and during
midnight  (from 23h  to  24h), in  which  \cosmoart perform  different
internal computations.  At  the same time we leverage  the two already
stated MPI policies: \emph{aggressive} and \emph{degraded}.

All  the  traces  in   Figure~\ref{fig:6}  are  divided  in  different
computation  phases. In first  phase all  the processes  perform group
communications      and       execute      MPI-receive      primitives
(e.g. \texttt{MPI\_Recv}). The second one performs computation bursts,
combined  in  between   with  group  communications,  blocking  sends,
immediate  receives and  MPI-wait primitives.  Finally,
group  communications interlaced  with computation  and  a significant
synchronization  phase are  performed. From  the performance  point of
view, one can notice a reduction of 2\,\% in the TTS that \emph{degraded}  MPI
mode leads over the \emph{aggressive} MPI  policy. On the other hand an averaged
reduction of power consumption of  2\,\% is attained with the \emph{degraded} 
MPI policy  only in  routines that actually  perform synchronizations,
group  communications  and  MPI-receive/-wait primitives.   Thus,  the
reduction of the power-energy  pair directly depends on the percentage
of time in which blocking MPI  operations are executed. For the 1 hour
simulation traces, an average of 50\,\% of the time the processes were
waiting for incoming messages.  Taking  into account that  the total power  reduction during
blocking phases  is about 5\,\%,  one can compute an  approximation of
the  power savings,  i.e., $50\,\%  \times  5\,\% =  2.5\%$, which  is
consistent  power  reduction  obtained  for  the  full-day  simulation
(around  2\,\%). However,  we  expect higher  power/energy savings  if
upcoming versions of OpenMPI support full blocking policies as well as if 
future architectures tend proportional computing~\citep{Barroso-2007}.

% \begin{figure*}[htbf]
%   \centering
%   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
%   Percentage of time for the 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
%   \caption{blabla}
%   \label{fig:7}
% \end{figure*}

### 2014-05-29 trunk/paper_enahpc14/paper.tex
cleans up latest version; improves format
Manuel's changes + rebuttal letter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\RequirePackage{fix-cm}
%
\documentclass[twocolumn]{svjour3}
%
\smartqed
%
\usepackage[numbers]{natbib}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{colortbl}
\usepackage{url}
\usepackage{xspace}

%
\DeclareRobustCommand\IPCClongname{}
\setcounter{secnumdepth}{5}
%
% please place your own definitions here and don't use \def but \newcommand{}{}
\newcommand{\pmlib}{\texttt{pmlib}\xspace}
\newcommand{\monch}{\textsc{Monch}\xspace}
\newcommand{\pilat}{\textsc{Pilatus}\xspace}
\newcommand{\tinto}{\textsc{Tintorrum}\xspace}
\newcommand{\cosmoart}{\textsc{Cosmo-Art}\xspace}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

@@

\begin{abstract}
  In this  paper we investigate  the energy footprint  and performance
  profiling  of  \textsc{COSMO-ART} on  various  HPC platforms.   This
  model is an  extension of the operational weather  forecast model of
  the German  weather service (DWD),  developed for the  evaluation of
  the interactions  of reactive gases  and aerosol particles  with the
  state of  atmosphere at  the regional scale.   Different measurement
  devices and  energy-aware techniques are described  to evaluate both
  time and  energy to  solution of the  considered application  and to
  gain detailed insights into power and performance requirements.  Our
  motivation  is to  improve  corresponding code  sections to  sustain
  performance  while minimizing energy-to-solution.   This preliminary
  work  sets the  basis for  subsequent studies  to  tackle challenges
  related  to  energy  efficient  high performance  computing  in  the
  framework of the Exa2Green project \cite{EXA2GREEN}.

\keywords{High performance computing  \and Energy-aware computing
  %Green computing  
  \and Numerical weather  prediction \and Atmospheric
  chemistry  \and  Aerosol  modelling \and
  Benchmark  analysis  \and 
  \cosmoart  coupled  model \and  
  Power-performance profiling and tracing tools  \and
  Energy-saving techniques}
% \PACS{PACS code1 \and PACS code2 \and more} 
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\RequirePackage{fix-cm}
%
\documentclass[twocolumn]{svjour3}
%
\smartqed
%
\usepackage[numbers]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{colortbl}
\usepackage{url}
\usepackage{xspace}
%
\DeclareRobustCommand\IPCClongname{}
%
% please place your own definitions here and don't use \def but \newcommand{}{}
\newcommand{\pmlib}{\texttt{pmlib}\xspace}
\newcommand{\monch}{\textsc{Monch}\xspace}
\newcommand{\pilat}{\textsc{Pilatus}\xspace}
\newcommand{\tinto}{\textsc{Tintorrum}\xspace}
\newcommand{\cosmoart}{\textsc{Cosmo-Art}\xspace}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

@@

\begin{abstract}
  In this  paper we investigate  the energy footprint  and performance
  profiling of \cosmoart  on various HPC platforms.  This  model is an
  extension of  the operational weather  forecast model of  the German
  weather  service   (DWD),  developed  for  the   evaluation  of  the
  interactions of reactive gases  and aerosol particles with the state
  of atmosphere at the  regional scale.  Different measurement devices
  and energy-aware techniques are  described to evaluate both time and
  energy  to  solution  of  the  considered application  and  to  gain
  detailed  insights  into power  and  performance requirements.   Our
  motivation  is to  improve  corresponding code  sections to  sustain
  performance  while minimizing energy-to-solution.   This preliminary
  work  sets the  basis for  subsequent studies  to  tackle challenges
  related  to  energy  efficient  high performance  computing  in  the
  framework of the Exa2Green project \cite{EXA2GREEN}.

\keywords{High performance computing  \and Energy-aware computing
  %Green computing  
  \and Numerical weather  prediction \and Atmospheric
  chemistry  \and  Aerosol  modeling \and
  Benchmark  analysis  \and 
  \cosmoart  coupled  model \and  
  Power-performance profiling and tracing tools  \and
  Energy-saving techniques}
% \PACS{PACS code1 \and PACS code2 \and more} 
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}
